---
title: "Functions in RAutoDiff"
author: Tore Selland Kleppe
output: rmarkdown::html_vignette
    
    
    
vignette: >
  %\VignetteIndexEntry{functions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(RAutoDiff)
```
# Introduction 
This document provides an overview of the different functions currently 
implemented (overloaded) for used with the Automatic Differentiation provided in
package `RAutoDiff`. Most of the functions are written in order to "work"
in the same way as the corresponding functions taking regular R-variables
as inputs. Therefore this document provides only limited description of these, and the documentation elsewhere is also rather sparse.

# Summary of functions and classes
## Classes
The package provides three S4 classes and one class union:

* classes `fAD` and `fAD2` represent numeric vectors (slot `val`), the Jacobian matrix (slot `jac`) of that vector with respect to the independent variables, and in the case of `fAD2`, an array of Hessian matrices (slot `hessian`, with respect to independent variables).
* classUnion `ADtype` is simply the union of `fAD` and `fAD2`.
* class `AD_matrix` is a matrix class were the matrix element values (slot `vals`) are stored as an `ADtype` (in row-major order). 

In what remains, `fAD` and `fAD2` should have a similar behavior (in the 
sense that the methods available for manipulating these should have the 
same names) as the built in numeric vector class `numeric`, and `AD_matrix` should have a similar behavior as built in numeric `matrix` class

Examples: (in general please avoid accessing and manipulating the slots of the classes directly)
```{r}
x.n <- c(1,2)
x <- independent2(x.n)
print(x@val) # value of variable
print(x@jac) # jacobian wrt to x.n
print(x@hessian) # hessians of x[i] with respect to x.n

m <- diag(x) # diagonal 2x2 AD_matrix
print(m@nrow)
print(m@ncol)
print(m@vals) 
```

In general, calculation with `fAD`-types is somewhat faster (close to fully vectorized) than for `fAD2` (several functions have for-loops), and hence `fAD2` should be avoided if you do not need second derivatives. 

## Creation and extraction

### Creating AD types

* `independent()` creates an `fAD` with values equal to the arguement and Jacobian equal to the identity
* `independent()` creates an `fAD2` with values equal to the arguement and Jacobian equal to the identity, and all Hessians equal to zero.

### Extracting from AD types

* `value()` Numeric value of the `ADtype` or `AD_matrix` (also overloaded as identity operations for `numeric` and `matrix`)
* `gradient()` Extract gradient of a scalar argument (length 1 `ADtype` or 1 by 1 `AD_matrix`)
* `jacobian()` Extract Jacobian of a vector (i.e. `ADtype`)
* `hessian()` Extract Hessian from a scalar `fAD2`

## Available functions for `ADtype`s

### Non-numerical

* Element access `[`,`[<-` e.g `x[1]<-x[2]` or `x[1:2] <- x[2]`
* `length()`
* Replications `rep.int()` and `rep_len()` e.g. `rep.int(x,4)`
* `show()`

### Binary numerical
All of the below apply also for mixed (`numeric` and `ADtype`) class argument, e.g. `3 + x` or `x * c(1,2.5)`. The methods recycle element in case of different sizes of the arguments.

* `+`, `-`, `*`, `/`, `^`

### Binary non-numerical

* `==`, `!=` `>`, `>=`, `<`, `<=`

### Unary numerical

* `+`, `-` e.g. `f <- -x`
* `exp()`, `log()`
* `sqrt()`
* `square()` (`square(x)` same as `x^2` but faster, also available for `numeric`)
* `lgamma()`
* `sin()`, `cos()`
* `abs()`
* To do:  more trigonometric, erf or pnorm, qnorm,cube

### Reductions

* `sum()`
* `prod()`
* To do: `min()`, `max()`, `pmax()`, `pmin()`

## Available functions for `AD_matrix`



### Non-numerical

* Element access `[`,`[<-` e.g `x[1,1]<-x[2,2]` or `x[1,] <- x[2]`
* `length()`, `nrow()`, `ncol()`
* `matrix()` where argument `data` is an `ADtype`
* `diag()` extract diagonal or create diagonal matrix
* `show()`
* `dim()`

### Elementwise binary numerical
Including between mixed types:

* `+`, `-`, `*`, `/`, `^`

### Elementwise unary numerical

* `+`, `-` 
* `exp()`, `log()`
* `sqrt()`
* `square()` 
* `lgamma()`
* `sin()`, `cos()`
* `abs()`

### Linear Algebra

* `%*%` for various combinations of arguments
* `backsolve()` and `forwardsolve()`
* `t()`
* `isSymmetric()`
* `solve()`
* `cholL()`, `chol()` and `solve.chol()` for SPD matrices

### Reductions

* `sum()`, `colSums()`, `rowSums()`
